---
title: "MovieLens Project Final Report"
author: "Khaliun Bat-Ochir"
date: "May 4th, 2021"
emails:
  - "khaliun83@yahoo.com"
  - "khaliun@spoon.mn"
linkedin: "https://www.linkedin.com/in/khaliun-bat-ochir-334925b4/"
info_program: "Data Science Professional Certificate Program"
info_institution: "HarvardX and online initiative by Harvard University"
info_course: "Capstone Course"
output:
  pdf_document:
    extra_dependencies: "subfig"
    toc: true
    toc_depth: 3
    number_sections: true
  word_document: default
papersize: a4
linestretch: 1.5
fontsize: 10pt
links-as-notes: true
bibliography: ["csl/references.bib"]
link-citations: yes
csl: "csl/harvard-the-university-of-sheffield-town-and-regional-planning.csl"
header-includes:
  - \usepackage{titling}
  - \pretitle{\begin{center}
    \includegraphics[width=2in,height=2in]{images/HarvardXLogo.jpeg}\LARGE\\}
  - \posttitle{\end{center}}
  - \usepackage{fancyhdr}
  - \pagestyle{fancy}
  - \fancyfoot[CO,CE]{Page Number}
  - \fancyfoot[LE,RO]{\thepage}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r dataprep-file-source, include=FALSE}
source("MovieLensAnalysisScripts.R", local = knitr::knit_global())
```

\newpage

# Overview

This is the `r rmarkdown::metadata$title` for `r rmarkdown::metadata$info_course` from `r rmarkdown::metadata$info_program` offered by `r rmarkdown::metadata$info_institution` prepared by `r rmarkdown::metadata$author`.

Final Report includes methodology and results of the project including RMSE calculation. Also, this report includes project environment description used for building and running the code.

## Project Goal

This is a Machine Learning project. Goal of the project consists in analysing **MovieLens 10M** and training data with chosen Machine Learning algorithm and reaching RMSE below 0.86490.

## Project Data

Project uses **MovieLens 10M** data set. The original code from provided by the instruction for Movielens Recommendation system project for Capstone Course downloads original data and as a [*Zip File*](http://files.grouplens.org/datasets/movielens/ml-10m.zip). Code then processes separate two files named **ratings** and **movies** included in the zip file that are joined together by "movieId" field.

## Project Files

Project files will be uploaded for review and grading by peers to Edx course section. Uploaded Files Include:
 
* RMarkdown Report File - "MovieLensAnalysisReport.Rmd" (Please note that this file sources MovieLensAnalysisScripts.R for running. You can find the code file in [*MovieLensCodes Github Repository*](https://github.com/khaliunb/MovieLensCodes.git)). Some alterations to the MovieLensAnalysisScripts.R are necessary if you should knit the report again.
* R Script File - "MovieLensAnalysisScripts.R" that includes codes and comments for model fitting and RMSE calculation for final model
* PDF Report - "MovieLensAnalysisReport.pdf"

For convenience purposes, total running time of each model fitting algorithm had been included in Result section of this report.
 
All the files have been also uploaded to [*MovieLensCodes Github Repository*](https://github.com/khaliunb/MovieLensCodes.git) 

## Project Environment

Codes for the project were built and tested using:

* R version 3.6 and
+ RStudio Version 1.3.1073
+ Linux Ubuntu 20.04

\newpage

# Method

## Original Code and Data Set Preparation

Original code divides downloaded data set into following two subsets:

* **edx** - equivalent of **training set** set that contains 80% of the complete **MovieLens10K** data set
 + **validation** - equivalent of **test set** that contains 20% of the complete **MovieLens10K** data set

Furthermore, **edx** data set had been matched with **validation** data set and all recurring data had been removed from **validation** data set.

Finally, the code removes temporary data sets (dl, ratings, movies, test_index, temp, movielens, removed) that were used to prepare **edx** and **validation** data sets.

Summary of **edx** data set:

```{r (edx)}
summary(edx%>%select(userId, movieId,rating, timestamp,title,genres))
```

Summary of **validation** data set:

```{r (validation), echo=FALSE}
summary(validation%>%select(userId, movieId,rating, timestamp,title,genres))
```

Note: Only **edx** data set have been used for data exploration, model fitting and training the data. **validation** data set will be used for final results and RMSE calculation.

## Data Exploration and Sampling method

### Linear model analysis data

Linear model analysis used for Machine Learning course will be applied for complete **edx** data set and results will be used for final model.

Regularized movie and user effect model have also been attempted on sample of 10'000.

However, the results were inconclusive as the sample data did not represent the completed **edx** data set. Therefore final *lambda* value for the model and analysis for the linear model have been carried out using complete **edx** data set.

**train_set** and **test_set** partitioned from complete **edx** data set had been used. 

* **train_set** - contains 80% of the complete **edx** data set
 + **test_set** - contains 20% of the complete **edx** data set

Furthermore, **train_set** data set had been matched with **test_set** data set and all recurring data had been removed from **validation** data set.

Summary of **train_set**:

```{r (train_set), echo=FALSE}
summary(train_set)
```

Summary of **test_set**:

```{r (test_set), echo=FALSE}
summary(test_set)
```

## Failures and Insights gained

Kmeans clustering have been tried for grouping movies into relevant groups as an attempt to replicate PCA analysis and incorporated the group ids to **edx** data set. However, the final RMSE for movie group model is 0.99 and does not go down for any values of k. Therefore the approach have been dropped. Code performing this analysis have been commented out in file * MovieAnalysisScripts_DataExploration.R *

## Model fitting and RMSE approach

Regularized movie and user effect model have be used for final model. For this, we will be using full cross-validation with **train_set** and **test_set** data sets for *lambda* parameter tuning. From there, minimum value of RMSE achieving *lambda* will be used final model training on **edx** data set.

For Final RMSE **validation** set will be used.

\newpage

# Results

Note: Results section contains plots and summary results present in code file *MovieAnalysisScripts_DataExploration.R* You can find relevant comments as a description for each result in this file.

## Linear model performance

We are getting a peek into full **edx** data set. For this, we are summarizing number of users and movies present in the data set.

```{r edx-user-movies-number, echo=FALSE}
edx %>%
  summarize(n_users = n_distinct(userId),
            n_movies = n_distinct(movieId))
```

```{r keep-prep-script, include=FALSE}
keep <- edx %>%
  dplyr::count(movieId) %>%
  top_n(5) %>%
  pull(movieId)
```

Now, let us list top 5 most rated movies in movielens data
```{r (keep)}
keep
```

```{r tab-prep-script, include=FALSE}
keep_raters<-edx %>% filter(movieId %in% keep) %>%
    mutate(rating=1) %>%
    select(movieId,userId,rating) %>%
    group_by(userId) %>%
    summarize(n=sum(rating)) %>%
    head()%>%
    pull(userId)

tab <- edx %>%
  filter(userId %in% keep_raters) %>% 
  filter(movieId %in% keep) %>%
  arrange(userId) %>%
  select(userId, title, rating) %>% 
  spread(title, rating)
```

Top 5 most rated movies' ratings in movielens data and transposes the title and rating columns by value and lists the results for userId column.
        
```{r (tab)}
tab %>% knitr::kable()
```

```{r ratings-spread-plot, fig.align='center', fig.cap='Imaging of the movies and users rating in 100 pixels', out.width="50%", echo=FALSE}
users <- sample(unique(edx$userId), 100)
par(mar = c(4, 4, .1, .1))
edx %>% filter(userId %in% users) %>% 
  select(userId, movieId, rating) %>%
  mutate(rating = 1) %>%
  spread(movieId, rating) %>% select(sample(ncol(.), 100)) %>% 
  as.matrix() %>% t(.) %>%
  image(1:100, 1:100,. , xlab="Movies", ylab="Users") 

abline(h=0:100+0.5, v=0:100+0.5, col = "grey")
```

We are showing the density plots for distribution of Movies and Users ratings.

```{r, fig.cap='Density Plots for Rating: (a) Movies (b) Users', fig.subcap=c('Movies Ratings Density', 'Users Ratings Density'), figures-side, fig.show="hold", out.width="50%", echo=FALSE}
par(mar = c(4, 4, .1, .1))
edx %>% 
  dplyr::count(movieId) %>% 
  ggplot(aes(n)) + 
  geom_histogram(bins = 30, color = "black") + 
  scale_x_log10() + 
  ggtitle("Movies")

edx %>%
  dplyr::count(userId) %>% 
  ggplot(aes(n)) + 
  geom_histogram(bins = 30, color = "black") + 
  scale_x_log10() +
  ggtitle("Users")
```
As we can see, the (b) Users Ratings Density plot differs from the Machine Learning course analysis results. Users are no longer rating mostly 1 time. This can be attributed to the fact that we are using more complete data set *MovieLens10M*.

### lambda value

We will remember that *lambda* is a tuning parameter. Therefore we will use cross-validation to choose it and apply the final minimum RMSE value of *lambda* for final training.

```{r linear-model-rmse, include=FALSE}
lambdas <- seq(0, 10, 0.25)
RMSE <- function(true_ratings, predicted_ratings){
  sqrt(mean((true_ratings - predicted_ratings)^2))
}
mu <- mean(train_set$rating)
rmses <- sapply(lambdas, function(l){
  b_i <- train_set %>%
    group_by(movieId) %>%
    summarize(b_i = sum(rating - mu)/(n()+l))
  b_u <- train_set %>% 
    left_join(b_i, by="movieId") %>%
    group_by(userId) %>%
    summarize(b_u = sum(rating - b_i - mu)/(n()+l))
  predicted_ratings <- 
    test_set %>% 
    left_join(b_i, by = "movieId") %>%
    left_join(b_u, by = "userId") %>%
    mutate(pred = mu + b_i + b_u) %>%
    .$pred
  return(RMSE(predicted_ratings, test_set$rating))
})

lambda <- lambdas[which.min(rmses)]
rmse_results <-data_frame(method="Regularized Movie + User Effect Model",
                                     RMSE = min(rmses))
```

```{r lambda-qplot, fig.align='center', fig.cap='Qplot: Lambda against RMSE', out.width="50%", echo=FALSE}
qplot(lambdas, rmses)
```

Analysis results for the Regularized Movie + User Effect model shows that the optimal value of lambda is as follows:

```{r lambda-optimal-value, echo=FALSE}
lambda
```
This value we will incorporate in the actual training of the linear model.

### Regularized Movie + User Effect model RMSE

Result for minimum RMSE gained from tuning lambda for Regularized Movie + User Effect is as follows:

```{r lambda-min-rmse-value, echo=FALSE}
rmse_results %>% knitr::kable()
```

## Final results, steps and timing of training on **edx** data set

We are using the *lambda* value gained using cross validation. We are incorporating the *mu*, *b_i*, *b_u*, *pred* features back into **edx** and **validation** data sets in file **MovieLensAnalysisScripts.R**. We will be only using *pred* feature for prediction and training Data preparation script runs approximately 2 minutes.

Then, we are fitting the model using *lm()* in file **MovieLensAnalysisScripts_RMSE_edx_dataset.R**.

Training of complete **edx** set was performed. Final RMSE is as follows:

```{r final-rmse-value, echo=FALSE}
final_rmse_results <-data_frame(data_frame(method="Regularized Movie + User Effect Model",
                          FINAL_RMSE = "0.8648617"))
final_rmse_results %>% knitr::kable()
```

Total time of execution took a second.

```{r timing-results, echo=FALSE}
timing_results <-data_frame(log=c("Data source prep script run time","Edx Data set training script run time"),
                            DURATION=c("132.39938 secs","1.17256 secs"))
timing_results %>% knitr::kable()
```

However, we should note that running time depends on the environment.

\newpage

# Conclusion

For final result, Regularized movie and user effect model have been used. Final RMSE is *0.8648617*. Achieved project goal of **RMSE < 0.86490**.

\newpage

# Contact Information {-}

If you have any questions regarding the project, please feel free to contact me at any of my emails: *`r rmarkdown::metadata$emails`*; or feel free to visit my [*Linkedin Profile*](`r rmarkdown::metadata$linkedin`)

# References {-}

## Books {-}

* @HarvardXDataScienceProgramTextbook
* @RMarkdownCookbook

## Articles {-}

* @kmeansclustering

## Manuals {-}

* @R-base
* @R-dplyr
* @R-ggplot2
* @R-tidyverse
* @lm